---
title: "Report SHINY PLOT ONLY"
output: html_document
runtime: shiny
---

# THIS ONLY HAS THE EDA SHINY INTERACTIVE PLOT THAT CANNOT RUN ON HTML

```{r, echo=FALSE}
suppressWarnings({
  suppressMessages({
  library(tidyverse)
  library(maps)
  library(lubridate)
  library(shiny)
  library(slider)
  library(ranger)
  library(pROC)
  library(rsconnect)
  })
})


```

```{r, results ='hide', echo=FALSE}
#dataset with 100k wildfires
wildFire100k <- read_csv(file = "./Data/wildfires_sample_100k.csv", show_col_types = FALSE) 
wildFire100k <- wildFire100k %>%
  filter(STATE != "AK", STATE != "HI", STATE != "PR", STATE != "DC")

```

```{r, echo=FALSE}
state_monthly_year_temp <- read_csv(file = "./Data/states_monthly_yearly_temp.csv", show_col_types = FALSE)
```

```{r, results='hide', echo=FALSE}
#Filtered for notable fires
notableFires <- wildFire100k %>% 
  select(FIRE_YEAR, DISCOVERY_DOY, FIRE_SIZE, LATITUDE, LONGITUDE, STATE) %>%
  mutate(STATE = as.factor(STATE)) %>%
  filter(FIRE_SIZE >= 300)
```

```{r, results='hide', echo=FALSE}
#Mutate column for FireDate and get Year and Month individually
notableFiresTime <- notableFires %>% 
    mutate(FIRE_DATE = as.Date(paste(FIRE_YEAR, DISCOVERY_DOY, sep = "-"), format = "%Y-%j"),
          YEAR = year(FIRE_DATE),
          MONTH = month(FIRE_DATE))
```

```{r, results='hide', echo=FALSE}
suppressMessages(
  #GRID and for each lat and long
notableFiresTime <- notableFiresTime %>%
  mutate(
    lat_bin = floor(LATITUDE),
    lon_bin = floor(LONGITUDE),
    ym = YEAR * 100 + MONTH   # e.g., 201507
  )
)

```

```{r, results='hide', echo=FALSE}
suppressMessages({
  # Fire indicator at (grid, month): did any notable fire occur?
 grid_month_fire <- notableFiresTime %>% 
  group_by(lat_bin, lon_bin, ym) %>%
  summarize(n_fires = n(),
            fire_this_month = as.numeric(n_fires > 0))

head(grid_month_fire)
})

```

```{r, results='hide', echo=FALSE}
# ---- Build panel (all grid cells that appear at least once) ----
all_grids <- grid_month_fire %>% distinct(lat_bin, lon_bin)

# Generate a safe monthly sequence using actual dates
start_date <- ymd(paste0(min(grid_month_fire$ym) %/% 100, "-", min(grid_month_fire$ym) %% 100, "-01"))
end_date   <- ymd(paste0(max(grid_month_fire$ym) %/% 100, "-", max(grid_month_fire$ym) %% 100, "-01"))

all_months <- tibble(date = seq.Date(start_date, end_date, by = "month")) %>%
  mutate(year = year(date), month = month(date), ym = year * 100 + month) %>%
  select(ym, year, month)

# Full panel = grid × month

panel <- crossing(all_grids, all_months) %>%
  left_join(grid_month_fire, by = c('lat_bin', 'lon_bin', 'ym') )

#Any NAs are to be replaced with 0 as that month, year, location did not have a notable wildfire
panel$n_fires[is.na(panel$n_fires)] <- 0
panel$fire_this_month[is.na(panel$fire_this_month)] <- 0
```

```{r, results='hide', echo=FALSE}
panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    y_next = lead(fire_this_month, n = 1L)
  ) %>%
  ungroup() %>%
  filter(!is.na(y_next))

table(panel$y_next)
```

```{r, results='hide', echo=FALSE}
suppressMessages({
  
  #function that picks the largest 
 mode_state <- function(x) { 
  uniqueVals <- unique(x)
  count <- tabulate(match(x, uniqueVals))
  uniqueVals[which.max(tabulate(match(x, uniqueVals)))] 
   
 }


 grid_state <- notableFiresTime %>% 
   group_by(lat_bin, lon_bin) %>%
   summarize(state = mode_state(STATE))

panel <- panel %>%
  left_join(grid_state, by = c("lat_bin","lon_bin"))

panel %>% select('state') %>% summary
  
})

```

```{r,  results='hide', echo=FALSE}
state_monthly_year_temp

state.nameNoAkHa <- state.name[state.name != 'Hawaii' & state.name != 'Alaska']
state.abbNoAkHa <- state.abb[state.abb != 'HI' & state.abb != 'AK']
forAbbtoFull <- setNames(tolower(state.nameNoAkHa), state.abbNoAkHa)
panel$state <- 
  forAbbtoFull[panel$state]

#turn month into a numeric form for both
panel$month <- as.numeric(panel$month)
state_monthly_year_temp$month <- as.numeric(state_monthly_year_temp$month)

#create rolling 3 month temperature average 
state_monthly_year_temp_prepped <- state_monthly_year_temp %>%
  arrange(state, year, month) %>%
  group_by(state) %>%
  mutate(
    'avg_temp_roll_3' = slide_dbl(average_temp, mean, .before = 2)
  )

#join so each state and date has a temperature
panel <- panel %>%
  left_join(state_monthly_year_temp_prepped, by = c("year", 'month', 'state')) 

```

```{r, results='hide', echo=FALSE}
# make a cos and sin column that calculates sin or cos of ((2 * pi) * (month/12))
 panel <- panel %>% 
  mutate('month_sin' = sin((2*pi) * (panel$month/12)),
         'month_cos' = cos((2*pi) * (panel$month/12)))
  
#########

panel %>% select('month_sin', 'month_cos') %>% head
```

```{r, results='hide', echo=FALSE}

suppressMessages({
    # mutate fire_size which will be used to calculate size_roll3 and size_roll12
grid_fire_size <- notableFiresTime %>%
  group_by(lat_bin, lon_bin, ym) %>%
  summarize(fire_size = sum(FIRE_SIZE))


panel <- panel %>%
  left_join(grid_fire_size, by = c("lat_bin","lon_bin", "ym")) %>%
  mutate(fire_size = coalesce(fire_size, 0))

#creates our new variables
panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>% 
  mutate(
    #sees if there was a wildfire a year ago in same location and month  ,
    'fire_lag12' = coalesce(lag(n_fires, 12), 0), 
    'fire_roll3' = slide_dbl(n_fires, sum, .before = 2), 
    'fire_roll12' = slide_dbl(n_fires, sum, .before = 11),

    'size_roll3' = 
      ifelse(fire_roll3 == 0, 0, slide_dbl(fire_size, sum, .before = 2)/fire_roll3),
    'size_roll12' = 
      ifelse(fire_roll12 == 0, 0, slide_dbl(fire_size, sum, .before = 11)/fire_roll12)
  )

panel %>% 
  select("n_fires", "fire_lag12", "fire_roll3", "fire_roll12", 
         "size_roll3", "size_roll12") %>%
  summary()


})


```

# 2) Exploratory Data Analysis

Lets analyze our dataset a little more:

```{r, results='hide', echo=FALSE}
#For Time and Space Interactive PLOT

#Set up eda_spatial for shiny plot and group by year
eda_spatial <- panel %>% 
  group_by(lat_bin, lon_bin, year) %>%
  mutate(fire_rate = mean(y_next))
```

```{r}
ui <- fluidPage(
   sliderInput("Year", "Select a Year from 1992 to 2015", 
               value = 1992, min = 1992, max = 2015),
  mainPanel(
  plotOutput("notableFireRate", height = "400")
)
)
```

```{r}
server <- function(input, output, session) {
  output$notableFireRate <- renderPlot({
    #
    filteredByYear <- eda_spatial %>%
      filter(year == input$Year[1]) 
    
    ggplot(filteredByYear, aes(x = lon_bin, y = lat_bin, fill = fire_rate)) +
  coord_quickmap(xlim = c(-125, -65), ylim = c(25, 50)) + 
  geom_tile() +
  labs(
    title = paste("Spatial and Time pattern: Mean next-month notable-fire rate by grid cell in", input$Year[1]),
    x = "Longitude bin (floor)",
    y = "Latitude bin (floor)",
    fill = "Mean y_next")
  })
}
shinyApp(ui, server)
```

This plot establishes the fact that **not all areas get a lot of "notable fires"** (or at all in a subset sample of 100,000 in this report) as some grid cells don't show up in the heat map. We also notice that **some areas consistently have fires every couple of years**. One notable observation I found is that an area (couple of grid cells in a cluster) that had a higher than normal mean `y_next` (any lighter shades of blue) on the graph tends to be completely dark the next year. For example, if we look at northern Texas on the plot in 2011, we will see a relatively large groupings of cells that are light shaded. The next year 2012, that area is almost completely dark. **Areas that had higher big wildfire rates last year tend to have lower large wildfire rates the following year**. **States, lat/long coordinates, and years** seem like they will be big factors to predicting the following month's potential for large wildfires.


```{r}
eda_month <- panel %>% 
  group_by(month, year) %>%
  mutate("mean_y_next" = mean(y_next))

```

```{r}
ui <- fluidPage(
   sliderInput("Year", "Select a Year from 1992 to 2015", 
               value = 1992, min = 1992, max = 2015),
  mainPanel(
  plotOutput("seasonFireRate", height = "400")
)
)
```

```{r}
server <- function(input, output, session) {
  output$seasonFireRate <- renderPlot({
    #
    filteredByYear <- eda_month %>%
      filter(year == input$Year[1]) 
    
  ggplot(filteredByYear, aes(x = month, y = mean_y_next)) +
    geom_line() +
    geom_point() +
    coord_cartesian(ylim = c(0, .06)) +
    scale_x_continuous(breaks = 1:12) +
    labs(
      title = paste("Seasonality: Mean probability of a notable fire in the next month in",
                    input$Year[1]),
      x = "Month",
      y = "Mean of y_next"
    )
  })
}
shinyApp(ui, server)
```

Looking at the seasonality plot, there is a **clear pattern of the summer months and sometimes late spring months like May showing higher averages of probabilities of a notable fire in the next month**. Interestingly, 2009 is somewhat of an anomaly by having its peak mean probability of a notable fire being in the winter months leading to spring (February and March). If we look back on 2008, we do notice that its summer months rates for notable fires are a lot lower than average which could indicate why 2009 seem to have an early burst of notable wildfires. I also want to mention 2011 which also had a notably high probability rate during the winter months which were accompanied by a 2010 which was overall a low probability year for large wildfires. **Months, and once again years, seem to be important**. Rolling monthly temperature could be important as well as it can detect the past few month's temperature and the current trend of the season/year(getting hotter vs getting colder).

