---
title: "Report"
output: html_document
---

```{r}
suppressMessages({
  library(tidyverse)
  library(maps)
  library(lubridate)
  library(shiny)
  library(slider)
  library(randomForest)
  library(pROC)
  
})

```

#Abstract


# 1) Data Preprocessing

## Datasets used

The full dataset [link](https://www.kaggle.com/datasets/rtatman/188-million-us-wildfires/data) contains 1.88 million entries of wildfires. The dataset of is of 795 MB in size. I use a smaller subset that contains 100,000 entries provided by Dr. Shizhe Chen for his Winter 2026 STA141A course at UC Davis. Detailed description of variables in the dataset is available on Kaggle.

```{r}
#dataset with 100k wildfires
wildFire100k <- read_csv(file = "./Data/wildfires_sample_100k.csv") 
```

I also used [link](https://www.kaggle.com/datasets/justinrwong/average-monthly-temperature-by-us-state) to contribute the monthly/annual temperature of the states to this project. Credit to Justin Wong.

What I modified:

-   Creating a new columns: `avg_yearly_temp` and `month_hotter_than_year_avg` :

    -   `avg_yearly_temp` - contains the average annual temperature of that state

    -   `month_hotter_than_year_avg` - compares and sees if `average_temp` is greater than `avg_yearly_temp`

-   include ONLY years between 1992 and 2015

-   removed all other columns outside of `month`, `year`, `state`, `average_temp`, and two self-created columns: `avg_yearly_temp` and `month_hotter_than_year_avg`

-   put all state names in column `state` into lowercase

```{r}
state_monthly_year_temp <- read_csv(file = "./Data/states_monthly_yearly_temp.csv")
```

## Temporal Patterns and Threshold for "Notable" Fires

For our predictive purposes, I want to emphasize and use these specific columns: - `FIRE_YEAR` - `DISCOVERY_DOY` - `FIRE_SIZE` - `LATITUDE` - `LONGITUDE` - `STATE` So I will mainly focus on these columns.

We use `FIRE_YEAR` and `DISCOVERY_DOY` to create a new column `FIRE_DATE` that puts the date in which the fire started in the format Y-M-D. We can also then separate that to get the individual year and month of each fire in the `YEAR` and `MONTH` column.

For our predictive tasks, we are interested in identifying and predicting "notable" fires that can cause major damage. In order to reflect this, the dataset is filtered to include only fires with a final burned size of 300 acres (`FIRE_SIZE` \>= 300).

```{r}
#Filtered for notable fires
notableFires <- wildFire100k %>% 
  select(FIRE_YEAR, DISCOVERY_DOY, FIRE_SIZE, LATITUDE, LONGITUDE, STATE) %>%
  mutate(STATE = as.factor(STATE)) %>%
  filter(FIRE_SIZE >= 300)
```

```{r}
#Mutate column for FireDate and get Year and Month individually
notableFiresTime <- notableFires %>% 
    mutate(FIRE_DATE = as.Date(paste(FIRE_YEAR, DISCOVERY_DOY, sep = "-"), format = "%Y-%j"),
          YEAR = year(FIRE_DATE),
          MONTH = month(FIRE_DATE))
```

## Creating a Grid for Spatial Pattern

```{r}
#GRID and for each lat and long
notableFiresTime <- notableFiresTime %>%
  mutate(
    lat_bin = floor(LATITUDE),
    lon_bin = floor(LONGITUDE),
    ym = YEAR * 100 + MONTH   # e.g., 201507
  )
```

```{r}
# Fire indicator at (grid, month): did any notable fire occur?
 grid_month_fire <- notableFiresTime %>% 
  group_by(lat_bin, lon_bin, ym) %>%
  summarize(n_fires = n(),
            fire_this_month = as.numeric(n_fires > 0))

head(grid_month_fire)
```

```{r}
# ---- Build panel (all grid cells that appear at least once) ----
all_grids <- grid_month_fire %>% distinct(lat_bin, lon_bin)

# Generate a safe monthly sequence using actual dates
start_date <- ymd(paste0(min(grid_month_fire$ym) %/% 100, "-", min(grid_month_fire$ym) %% 100, "-01"))
end_date   <- ymd(paste0(max(grid_month_fire$ym) %/% 100, "-", max(grid_month_fire$ym) %% 100, "-01"))

all_months <- tibble(date = seq.Date(start_date, end_date, by = "month")) %>%
  mutate(year = year(date), month = month(date), ym = year * 100 + month) %>%
  select(ym, year, month)

# Full panel = grid × month

panel <- crossing(all_grids, all_months) %>%
  left_join(grid_month_fire, by = c('lat_bin', 'lon_bin', 'ym') )

#Any NAs are to be replaced with 0 as that month, year, location did not have a notable wildfire
panel$n_fires[is.na(panel$n_fires)] <- 0
panel$fire_this_month[is.na(panel$fire_this_month)] <- 0
```

## Predictability Column

We add a binary column `y_next` that returns 1 if the same lat/long cell next month has at least one notable wildfire. (0 is not)

```{r}
panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    y_next = lead(fire_this_month, n = 1L)
  ) %>%
  ungroup() %>%
  filter(!is.na(y_next))

table(panel$y_next)
```

## Baseline Predictors

### `state`

This [link](https://www.statology.org/mode-in-r/) helped me develop the `mode_state` function for determining which state appears the most frequent in terms of notable wildfire frequency.

```{r}
#function that picks the largest 
 mode_state <- function(x) { 
  uniqueVals <- unique(x)
  count <- tabulate(match(x, uniqueVals))
  uniqueVals[which.max(tabulate(match(x, uniqueVals)))] 
   
 }


 grid_state <- notableFiresTime %>% 
   group_by(lat_bin, lon_bin) %>%
   summarize(state = mode_state(STATE))

panel <- panel %>%
  left_join(grid_state, by = c("lat_bin","lon_bin"))

panel %>% select('state') %>% summary
```

### Seasonality (MONTHS)
We want to make it feel more like a cycle as numerically January and December are far apart; however, in reality they are close together.

```{r}
# make a cos and sin column that calculates sin or cos of ((2 * pi) * (month/12))
 panel <- panel %>% 
  mutate('month_sin' = sin((2*pi) * (panel$month/12)),
         'month_cos' = cos((2*pi) * (panel$month/12)))
  
#########

panel %>% select('month_sin', 'month_cos') %>% head
```
### Lagged fire history
`n_fires` - # of fires occurred that month and grid
`fire_lag12` - # of fires occurred that month and grid one year ago
`fire_roll3` - # of fire that occurred in this grid cell the last three months
`fire_roll12` - # of fire that occurred in this grid cell the last year
`size_roll3` - average fire size in this grid cell the last three months
`size_roll12` - average fire size in this grid cell the last year

```{r}
# mutate fire_size which will be used to calculate size_roll3 and size_roll12
grid_fire_size <- notableFiresTime %>%
  group_by(lat_bin, lon_bin, ym) %>%
  summarize(fire_size = sum(FIRE_SIZE))


panel <- panel %>%
  left_join(grid_fire_size, by = c("lat_bin","lon_bin", "ym")) %>%
  mutate(fire_size = coalesce(fire_size, 0))

#creates our new variables
panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>% 
  mutate(
    #sees if there was a wildfire a year ago in same location and month  ,
    'fire_lag12' = coalesce(lag(n_fires, 12), 0), 
    'fire_roll3' = slide_dbl(n_fires, sum, .before = 2), 
    'fire_roll12' = slide_dbl(n_fires, sum, .before = 11),

    
    'size_roll3' = 
      ifelse(fire_roll3 == 0, 0, slide_dbl(fire_size, sum, .before = 2)/fire_roll3),
    'size_roll12' = 
      ifelse(fire_roll12 == 0, 0, slide_dbl(fire_size, sum, .before = 11)/fire_roll12)
  )
#ifelse(fire_roll3 == 0, 0, slide_dbl(fire_size, sum, .before = 2)/fire_roll3
#ifelse(fire_roll12 == 0, 0, slide_dbl(fire_size, sum, .before = 11)/fire_roll12)
panel %>% 
  select("n_fires", "fire_lag12", "fire_roll3", "fire_roll12", 
         "size_roll3", "size_roll12") %>%
  summary()

```

```{r}
state_monthly_year_temp

panel %>%
  left_join(state_monthly_year_temp)

panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    month_next_hotter = lead(month_hotter_than_year_avg, n = 1L)
  ) %>%
  ungroup() %>%
  filter(!is.na(y_next))
```

```{r}
#dataset_summary table
dataset_summary <- tibble(
  n_rows = nrow(panel),
  year_min = min(panel$year),
  year_max = max(panel$year),
  spatial_unit = "1° x 1° lat-lon grid (floor bins)",
  temporal_unit = "monthly",
  outcome = "y_next = 1 if any notable fire occurs in grid cell next month",
  predictors = "lat_bin, lon_bin, state, month_cos, month_sin, year, 
    n_fires, fire_lag12, fire_roll3, fire_roll12, 
    size_roll3, size_roll12"
)

```



# 2) Exploratory Data Analysis

```{r}
#For Time and Space Interactive PLOT

#Set up eda_spatial for shiny plot and group by year
eda_spatial <- panel %>% 
  group_by(lat_bin, lon_bin, year) %>%
  mutate(fire_rate = mean(y_next))

ui <- fluidPage(
   sliderInput("Year", "Select a Year from 1992 to 2015", 
               value = 1992, min = 1992, max = 2015),
  mainPanel(
  plotOutput("notableFireRate", height = "400")
)
)

server <- function(input, output, session) {
  output$notableFireRate <- renderPlot({
    #
    filteredByYear <- eda_spatial %>%
      filter(year == input$Year[1]) 
    
    ggplot(filteredByYear, aes(x = lon_bin, y = lat_bin, fill = fire_rate)) +
  coord_quickmap(xlim = c(-125, -65), ylim = c(25, 50)) + 
  geom_tile() +
  labs(
    title = paste("Spatial and Time pattern: Mean next-month notable-fire rate by grid cell in", input$Year[1]),
    x = "Longitude bin (floor)",
    y = "Latitude bin (floor)",
    fill = "Mean y_next")
  })
}
shinyApp(ui, server)
```

```{r}
eda_month <- panel %>% 
  group_by(month, year) %>%
  mutate("mean_y_next" = mean(y_next))


#########
ui <- fluidPage(
   sliderInput("Year", "Select a Year from 1992 to 2015", 
               value = 1992, min = 1992, max = 2015),
  mainPanel(
  plotOutput("seasonFireRate", height = "400")
)
)

server <- function(input, output, session) {
  output$seasonFireRate <- renderPlot({
    #
    filteredByYear <- eda_month %>%
      filter(year == input$Year[1]) 
    
  ggplot(filteredByYear, aes(x = month, y = mean_y_next)) +
    geom_line() +
    geom_point() +
    coord_cartesian(ylim = c(0, .06)) +
    scale_x_continuous(breaks = 1:12) +
    labs(
      title = paste("Seasonality: Mean probability of a notable fire in the next month in",
                    input$Year[1]),
      x = "Month",
      y = "Mean of y_next"
    )
  })
}
shinyApp(ui, server)
```

# 3) Modeling 

```{r}
model_df <- panel %>%
  mutate(y_next = factor(y_next, levels = c(0, 1))) %>%
  select(y_next, lat_bin, lon_bin, state, 
         month_cos, month_sin, year, 
         n_fires, fire_lag12, fire_roll3, fire_roll12, size_roll3, size_roll12)
glimpse(model_df)
```

In this report, we will fit two models:

1) Logistic Regression using the predictors: 
`lat_bin`, `lon_bin`, `state`, `month_cos`, `month_sin`, `year`, `n_fires`, `fire_lag12`, `fire_roll3`, `fire_roll12`, `size_roll3`, `size_roll12`
to predict `y_next` which indicates whether or not the next month will contain at least one "notable" wildfire. This model will serve as the benchmark for performance.



2) Random Forest with `ranger` package using the predictors: 
`lat_bin`, `lon_bin`, `state`, `month_cos`, `month_sin`, `year`, `n_fires`, `fire_lag12`, `fire_roll3`, `fire_roll12`, `size_roll3`, `size_roll12`
to predict `y_next`. A decision tree like random forest is able to handle non-linear trends between our predictors and `y_next` better than a logistic regression and it remains interpretable.

# 4 Model Fitting and Validation

[link](https://stackoverflow.com/questions/64060369/ranger-predict-incorrect-number-of-dimensions-in-r) helped improved my code for the random forest tree.
```{r}
years_sorted <- sort(unique(model_df$year))
min_train_years <- 3

#for random forest put response variable in factor 
feat_formula <- as.factor(y_next) ~ lat_bin + lon_bin + state + month_cos + month_sin + year +
  n_fires + fire_lag12 + fire_roll3 + fire_roll12 + size_roll3 + size_roll12

fold_results <- tibble(
    train_end_year = years_sorted[(min_train_years):(length(years_sorted) - 2)],
    valid_year = years_sorted[(min_train_years + 1):(length(years_sorted) - 1)]) %>%
  mutate(
    # Logistic Regression
      auc_glm = map2_dbl(train_end_year, valid_year, function(tr_end, va_year) {
          train <- model_df %>% filter(year <= tr_end)
          valid <- model_df %>% filter(year == va_year)
          
          ######### FILL IN THE MISSING PARTS #########
          fit <- glm(feat_formula , family = binomial, data = train)
          p_hat <- predict(fit, newdata = valid, type = "response")

          #########

          roc_obj <- roc(response = valid$y_next, predictor = p_hat, quiet = TRUE)
          as.numeric(auc(roc_obj))}),
      #Random Forest
    auc_rf = map2_dbl(train_end_year, valid_year, function(tr_end, va_year) {
      
      train <- model_df %>% filter(year <= tr_end)
      valid <- model_df %>% filter(year == va_year)
      fit <- ranger(feat_formula, data = train, num.trees = 200, mtry = 2,
                    probability = TRUE)

      p_rf <-predict(fit, data = valid)$predictions[, "1"]
      roc_obj <- roc(response = valid$y_next, predictor = p_rf, quiet = TRUE)
          as.numeric(auc(roc_obj))
      
      })
  )

fold_results

fit <- ranger(feat_formula, data = train, num.trees = 200, mtry = 2,
                    probability = TRUE)
predict(fit, data = valid)$predictions
```

```{r}

                        
      

      train_end_year = years_sorted[(min_train_years):(length(years_sorted) - 2)]
      valid_year = years_sorted[(min_train_years + 1):(length(years_sorted) - 1)]
     
      xgBoost_fit <- xgboost(x =  train[, -1], y = train$y_next)
      predict(xgBoost_fit, newdata = valid)

      train <- model_df %>% filter(year <= 2010)
          valid <- model_df %>% filter(year == 2011)
          as.matrix(train[, -1])

          xgBoost_fit <- xgboost(x = train_pred, y = train_response)
          xg_prob <- predict(xgBoost_fit, valid_pred)


          roc_obj <- roc(response = valid$y_next, predictor = xg_prob, quiet = TRUE)
          as.numeric(auc(roc_obj))})
      
      

  '        
          dtrain <- xgb.DMatrix(data = train_pred, label = train_response)

          
          param <- xgb.params(
            max_depth = 2,
            objective = "binary:logistic",
            eval_metric = "auc"
          )
          
          bst <- xgb.train(param, dtrain, nrounds = 2, verbose = 0)
          xg_prob <- predict(xgBoost_fit, valid_pred)
  '

```
